---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load the data and view the summary of the data:
```{r}
data <- read.csv(paste(getwd(),"/activity.csv", sep=""), stringsAsFactors=FALSE)
summary(data)
str(data)
```

Use lubridate to convert date variable from class 'chr' to class 'POSIXct':
```{r}
library(lubridate)
data$date <- ymd(data$date)
class(data$date)
```

## What is mean total number of steps taken per day?

Calculate the number of steps per day:
```{r}
library(dplyr)
byday <- data %>% group_by(date) %>% summarize(total=sum(steps, na.rm=TRUE))
```

Calculate the mean number of steps per day:
```{r}
mean <- mean(byday$total)
mean
```

Plot a histogram of the total number of steps per day using base plotting:
```{r}
hist(byday$total, main = "Histogram of Steps Taken Per Day", xlab = "Number of Steps", ylab = "Number of Days")
```

Calculate the mean and median daily number of steps:
```{r}
dailymean <- mean(byday$total)
dailymedian <- median(byday$total)
paste("Mean daily number of steps is",round(dailymean))
paste("Median daily number of steps is",round(dailymedian))
```

## What is the average daily activity pattern?

Calculate the average number of steps per interval across all days:
```{r}
byinterval <- data %>% group_by(interval) %>% summarize(mean=mean(steps, na.rm=TRUE))
```

Plot the time series of the average number of steps per interval:
```{r}
plot(byinterval$interval, byinterval$mean, type="l", main = "Time Series of The Average Number of Steps By Interval", xlab = "5-Minute Interval (hhmm)", ylab = "Average Number of Steps")
```

Calculate which interval contains the maximum average number of steps:
```{r}
Maxinterval <- byinterval[which(byinterval$mean == max(byinterval$mean)),1]
paste("The interval which contains the maximum average number of steps is", Maxinterval)
```

## Imputing missing values

Calculate total number of rows with 'NA':
```{r}
paste("Total number of rows with 'NA' is", length(data$steps[is.na(data$steps)]))
```

Fill in the missing values in the steps variable, using the mean for that 5-minute interval:
Step 1. Divide the data into those with 'NA' and those without:
```{r}
data$id <- 1:nrow(data)
data1 <- filter(data, is.na(steps))
dim(data1)
data2 <- filter(data, !is.na(steps))
dim(data2)
dim(data1) + dim(data2)
```

Step 2. Using the 'byinterval' data, fill in the steps variable that are 'NA' with the the mean steps for that interval:
```{r}
m <- merge(byinterval, data1, by="interval")
m <- mutate(m, steps = mean) %>% select(-2)
```

Step 3. Combine the split data back into one set:
```{r}
datafilled <- rbind(data2, m)
sum(is.na(datafilled))
datafilled <- arrange(datafilled, id) %>% select(-4)
```

Make a histogram of the average number of steps per day, calculate the mean and median, and compare to earlier unfilled data:
```{r}
bydayfilled <- datafilled %>% group_by(date) %>% summarize(total=sum(steps))
hist(bydayfilled$total, main = "Histogram of Steps Taken Per Day", xlab = "Number of Steps", ylab = "Number of Days")
dailymeanfilled <- mean(bydayfilled$total)
dailymedianfilled <- median(bydayfilled$total)
paste("Mean daily number of steps is",round(dailymeanfilled))
paste("Median daily number of steps is",round(dailymedianfilled))
diffmean <- dailymeanfilled - dailymean
diffmedian <- dailymedianfilled - dailymedian
paste("Difference in mean between filled and unfilled data is", round(diffmean))
paste("Difference in median between filled and unfilled data is", round(diffmedian))
```

## Are there differences in activity patterns between weekdays and weekends?

Create factor and factor labels, "weekday" and "weekend":
```{r}
whatday <- wday(datafilled$date)
daytype <- c(whatday == 1 | whatday == 7)
datafilled$daytype <- factor(daytype, labels=c("weekday", "weekend"))
```

Average across days:
```{r}
library(lattice)
byinterval1 <- datafilled %>% group_by(interval, daytype) %>% summarize(mean=mean(steps, na.rm=TRUE))
xyplot(mean~interval | daytype, data=byinterval1, layout = c(1,2), type = "l", main = "Average Number of Steps by Interval and Day Type", xlab = "Interval (hhmm)", ylab = "Average Number of Steps")
```
